---
title: "Lab 3 - Data Types | Exploring Datasets"
author: "Environmental Data Analytics | John Fay and Luana Lima "
date: "Spring 2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Objectives
1. Discuss and navigate different data types in R
2. Create, manipulate, and explore datasets
3. Date objects

## Data Types in R
R treats objects differently based on their characteristics. For more information, please see: https://www.statmethods.net/input/datatypes.html. 

* **Vectors** 1 dimensional structure that contains elements of the same type.
#one dimemsional sturcture, a list of elements 

* **Matrices** 2 dimensional structure that contains elements of the same type.
#two dimenstional struction 

* **Arrays** Similar to matrices, but can have more than 2 dimensions. We will not delve into arrays in depth.
#an element that can be analyzed by multiople dimentions 

* **Lists** Ordered collection of elements that can have different modes.


* **Data Frames** 2 dimensional structure that is more general than a matrix. Columns can have different modes (e.g., numeric and factor). When we import csv files into the R workspace, they will enter as data frames.
#different modes, there is a lot more freedom with data frames. They need to have all colomns with same lengths and if you don't know the data you must fill with NA 

Define what each new piece of syntax does below (i.e., fill in blank comments). Note that the R chunk has been divided into sections (# at beginning of line, ---- at end)
```{r}
# Vectors ----
vector1 <- c(1,2,5.3,6,-2,4) # numeric vector
vector1
vector2 <- c("one","two","three") # character vector #you have to use the quotes so that r knows what to name the values 
vector2
vector3 <- c(TRUE,TRUE,TRUE,FALSE,TRUE,FALSE) #logical vector #vectors with true or false, you don't need quotes but they need to be capital 
vector3

vector1[3] # 
#if you want to find an element is another vector

# Matrices ----
matrix1 <- matrix(1:20, nrow = 5,ncol = 4) # you need to specify number of rows and columns, it fills out all of column one first and then moves to the next row
matrix1
matrix2 <- matrix(1:20, nrow = 5, ncol = 4, byrow = TRUE) #this will fill the data by row 
matrix2
matrix3 <- matrix(1:20, nrow = 5, ncol = 4, byrow = TRUE, # return after comma continues the line
                  dimnames = list(c("uno", "dos", "tres", "cuatro", "cinco"), 
<<<<<<< HEAD
                                 c("un", "deux", "trois", "cat"))) #
matrix3 
=======
                                  c("un", "deux", "trois", "cat"))) #
matrix3
>>>>>>> b3c7106116c07efb5453f2e1b6ba3c74dfab9fdd

matrix1[4, ] #we are just looking for row 4 and want all columns 
matrix1[ , 3] #we are just looking for column 3 in maxtrix 1 
matrix1[c(12, 14)] #
matrix1[c(12:14)] #
matrix1[2:4, 1:3] #

cells <- c(1, 26, 24, 68)
rnames <- c("R1", "R2")
cnames <- c("C1", "C2") 
matrix4 <- matrix(cells, nrow = 2, ncol = 2, byrow = TRUE,
  dimnames = list(rnames, cnames)) # 
matrix4

# Lists ---- 
list1 <- list(name = "Maria", mynumbers = vector1, mymatrix = matrix1, age = 5.3); list1
list1[[2]]

# Data Frames ----

dataframe1$ID


```
Question: How do the different types of data appear in the Environment tab?

> Answer: Data, values, function 

Question: In the R chunk below, write "dataframe1$". Press `tab` after you type the dollar sign. What happens?

> Answer: The teab show you need help finding the name of the object, you can do this in the console, and when you start typing in the counsole it will start to give suggestions  




### Coding challenge

Find a ten-day forecast of temperatures (Fahrenheit) for Durham, North Carolina. Create two vectors, one representing the high temperature on each of the ten days and one representing the low.

```{r}
high_temp <- c(49, 58, 39, 41, 44, 37, 41, 51, 52, 52)
high_temp
low_temp <- c(30, 30, 18, 26, 26, 18, 26, 26, 26, 28 )
low_temp

```

Now, create two additional vectors that include the ten-day forecast for the high and low temperatures in Celsius. Use a function to create the two new vectors from your existing ones in Fahrenheit.

```{r}
fahrenhiet.to.celsius <- function(high_temp) {
  (high_temp-32/1.8)
}
fahrenhiet.to.celsius(high_temp)
fahrenhiet.to.celsius(low_temp)

```

Combine your four vectors into a data frame and add informative column names.

```{r}
#data_frame
#durham_temp 
high_temp
low_temp
Durham_weather <- data.frame("Durham High Temp"=high_temp, "Durham Low Temp"=low_temp)
Durham_weather
```

Use the common functions `summary` and `sd` to obtain basic data summaries of the ten-day forecast. How would you call these functions differently for the entire data frame vs. a single column? Attempt to demonstrate both options below.

```{r}

```

### Date objects

Remember formatting of dates in R: 

%d  day as number (0-31)
%m  month (00-12, can be e.g., 01 or 1)
%y  2-digit year
%Y  4-digit year
%a  abbreviated weekday
%A  unabbreviated weekday
%b  abbreviated month
%B  unabbreviated month

```{r}
# Adjust date formatting for today
# Write code for three different date formats. 
# An example is provided to get you started.
# (code must be uncommented)
today <- Sys.Date()
format(today, format = "%B")
format(today, format = "%y")
format(today, format = "%m-%d-%Y")
#format(today, format = "")

```


### Package lubridate

Install and load the package lubridate into your R session. Lubridate offers fast and user friendly parsing of date-time data. Create a string for today's data and then convert it to R date object using lubridate.

More info on lubridate [here][https://cran.r-project.org/web/packages/lubridate/lubridate.pdf].

```{r DateExercise1}
install.packages("lubridate")
library(lubridate)

#Ex1
str_today <- "2021-feb-3"
#Since the format is year-month-day we will use function ymd()
date_obj_today <- ymd(str_today)
date_obj_today
#funtion to make the character of the date into a function

#Ex2
str_today <- "21-feb-3"
#Sine the format is year-month-day we will use function ymd()
date_obj_today <- ymd(str_today)
date_obj_today

#there are other similar functions ydm(), mdy(), etc
```

```{r DateExercise2}
#century issue
str_past <- "55-feb-3"
date_obj_past <- ymd(str_past)
date_obj_past

#Build a function to fix year that is more general than the one discussed in the lesson
fix.early.dates <- function(d, cutoff) {
       m <- year(d) %% 100  #operator %% is a modular division i.e. integer-divide year(d) by 100 and returns the remainder 
       year(d) <- ifelse(m > cutoff, 1900+m, 2000+m)  #this will update year(d), year() is a function that returns the year for a data object
       return(d)
}

fixed_date_obj_past <- fix.early.dates(date_obj_past,cutoff=21) #cutoff could be the current year to be more general or any other depending on data set 
fixed_date_obj_past
```

```{r centuryfix}
#Fix for century issue
str_past <- "55-feb-3"
#Alternative 1
date_obj_past <- fast_strptime(str_past,"%y-%b-%d",cutoff_2000=22L)
date_obj_past

#Alternative 2
date_obj_past2 <- parse_date_time2(str_past,"ymd",cutoff_2000=22L)
date_obj_past2

#Functions ymd(), mdy(), ydm() do not take argument cutoff_2000
```

In some cases when dates are provided as integers, you may need to provide an origin for your dates. For example, excel date could be given as number of days since an origin date. Origin date can be different. When R looks at dates as integers, its origin is January 1, 1970. Check if that is true on your machine. 

```{r dateorigin}
#Check if "1970-01-01" is your origin date.
lubridate::origin

```