---
title: "Assignment 2: Coding Basics"
author: Diane Sanchez
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your first and last name into the file name (e.g., "FirstLast_A02_CodingBasics.Rmd") prior to submission.



## Basics Day 1

1. Generate a sequence of numbers from one to 100, increasing by fours. Assign this sequence a name. 


2. Compute the mean and median of this sequence. 


3. Ask R to determine whether the mean is greater than the median. 

4. Insert comments in your code to describe what you are doing.


```{r}
#1.
# this is a sequence of Green Devil responses to survey.  
response_rate_GD <- seq( 1, 100, 4) # from, to, by 
seq( 1, 100, 4) # from, to, by 

#2. 

mean(response_rate_GD)
median(response_rate_GD)
#I am running summary stats data on response rates

#3. 
mean(response_rate_GD)> median(response_rate_GD)
#this is a simple comparison of summary stats. 

```

## Basics Day 2

5. Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6. Label each vector with a comment on what type of vector it is.

7. Combine each of the vectors into a data frame. Assign the data frame an informative name.

8. Label the columns of your data frame with informative titles.

```{r} 
#data_frame 
#student_test_score
student <- c ('Sam', 'LD', 'Phil', 'Abbey', 'Jarius', 'Geroldine')
Score <- c ( 100 , 90 , 80 , 70 , 60 , 40 )
passed_test <- c( TRUE, TRUE, TRUE, TRUE, TRUE, FALSE)

student_test_score <- data.frame("Names"=student,"TestScores"= Score,"Test Results"=passed_test)
student_test_score

#this is a data frame of studnet test results and whether they passed or not. 

```

9. QUESTION: How is this data frame different from a matrix?

> Answer: A data frame allows for multiple forms of data to be analyzed, while a matrix is used to analyze same type of data. For example question 8 would only work with a data frame becuase it is a combination of diffent vectors (names, test scores, and if students passed), the matrix could not do that. 

10. Create a function with an if/else statement. Your function should determine whether a test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. Hint: Use `print`, not `return`. The name of your function should be informative.

11. Apply your function to the vector with test scores that you created in number 5. 

```{r}

Pass_failure <- function(x) {
   if (x>50) {
return (TRUE) 
} 
else {
return (FALSE)
}
  }
Pass_failure (student_test_score$TestScores)

Pass_fail <- function(x) {
  ifelse (x >= 50, TRUE, FALSE) 
  }
Pass_fail (student_test_score$TestScores)

#this is a function of whether the studnet passed their test or not.  
```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

>  Answer: the if else function can only it process one value at a time. As expalined in the erros message the if statement will only run on the first element, so in my case since the first value on my data frame was large than 50, the restult was true. 
ifelse worked becuase it goes through all the values in the vector and will assign either true or false based on what the data values your assigned. 
